Genome-wide mapping of Cas9-induced sister chromatid exchange across single and over 200 genomic targets

Version 1

Abstract

CRISPR/Cas9-induced DNA double-strand breaks (DSBs) are widely used for genome engineering, yet their capacity to provoke sister chromatid exchange (SCE) and associated genome instability remains incompletely understood, in part because exchanges between identical sister chromatids leave no sequence change and are invisible to conventional whole-genome sequencing. Using sci-L3-Strand-seq, a scalable single-cell template-strand sequencing platform for genome-wide SCE mapping, we quantified strand-switch events following targeted Cas9 cleavage. A single Cas9 cut at a unique genomic locus led to strong local enrichment of SCE at the break site, reaching up to 41% in the same cell cycle and 17% in the subsequent division, indicating that DSB repair frequently engages non-local inter-sister repair. Extending the assay to 237 repetitive targets revealed mild and marginally statistically significant enrichment of on-target SCE across wild-type Cas9 and nickase variants (D10A and H840A). However, restricting analysis to a subset of cells with elevated SCE burden (>8 SCEs per cell) uncovered significant enrichment at programmed cut sites, suggesting that recombination at repetitive loci is conditionally engaged in highly recombinogenic cells. Reciprocal daughter-cell pair analysis further revealed large-scale structural alterations on chromosomes with induced SCEs and at SCE junctions, indicating that Cas9-induced SCEs do not universally reflect error-free homologous recombination. Disruption of DNA repair genes at the cut site, including LIG3, LIG4, XRCC1, and XRCC4, did not measurably alter SCE frequency per cell, consistent with delayed functional loss following editing and, in the case of essential genes such as LIG3, selection against disruptive alleles. Together, these findings demonstrate that Cas9-induced DSBs are potent local triggers of SCE at unique loci, while at repetitive regions SCE is more context-dependent and can be associated with structural alterations, highlighting the influence of lesion type and genomic context on recombination outcomes during genome editing.

Introduction

CRISPR/Cas9 genome editing relies on the induction of site-specific DNA double-strand breaks (DSBs) to trigger endogenous repair pathways (Cong et al., 2013; Ran et al., 2013). The two principal repair outcomes, non-homologous end joining (NHEJ) and homologous recombination (HR), have been extensively characterized at the sequence level in wild-type cells and with DNA repair gene perturbations (Chen et al., 2019; Hussmann et al., 2021; Shen et al., 2018): NHEJ generates small insertions and deletions at the break site, while HR can incorporate an exogenous donor template for precise editing (Richardson et al., 2016). However, this framework captures only the local and mutational fraction of the repair landscape. For non-local mutational outcomes, the structural consequences of Cas9 cutting extend well beyond the small indels detected by standard amplicon sequencing. Large deletions spanning kilobases have been documented at on-target sites (Kosicki et al., 2018), and Cas9-induced DSBs can trigger chromothripsis through micronucleus formation and chromosome bridge breakage (Leibowitz et al., 2021). Chromosome loss and segmental aneuploidies have been observed following Cas9 editing in human embryos, attributed to the isochromatid nature of the break (Zuccaro et al., 2020). At the cytogenetic level, chromosomal instability and rearrangements in edited cancer cell lines are frequently missed by PCR-based genotyping (Regan et al., 2025). These observations establish that Cas9 DSBs can produce a spectrum of structural alterations, yet the relationship between SCE and these larger-scale events has not been examined in the same experimental framework.

In this study, we assay what is likely the most frequent non-local outcome induced by Cas9: sister chromatid exchange (SCE). Here, we operationally define SCE as a reciprocal strand switch between sister chromatids, without implying a single mechanistic pathway. SCE can therefore arise from inter-sister homologous recombination (HR), double non-homologous end joining (“dNHEJ”) across two isochromatid DSBs, or other non-local repair processes. This distinction is particularly relevant for Cas9, which typically cleaves both sister chromatids at the same locus and thereby generates the precise isochromatid substrate on which dNHEJ can act. If only one sister is cleaved, the result is a two-ended, post-replicative DSB with an intact sister available as a repair template, an ideal substrate for HR-mediated SCE. More commonly, however, Cas9 can generate isochromatid DSBs through two routes: by cleaving both post-replicative sisters independently, or by cutting the unreplicated template ahead of the fork, such that the break is replicated into DSBs on both daughters. In either case, four broken ends are produced and neither sister is intact (Zuccaro et al., 2020). The repair outcome then depends on how the four ends are resolved: cis joining (each sister re-ligated to its own ends) restores the original configuration, while trans joining between sisters (the proximal end of one sister joined to the distal end of the other) produces a strand switch operationally indistinguishable from HR-mediated SCE. Whether Cas9-induced SCE at isochromatid breaks arises through HR, through trans end joining (dNHEJ), or through a mixture of both has not been determined. In all cases, exchanges between identical sisters are genetically silent and invisible to conventional whole-genome sequencing, and have therefore been largely overlooked as an outcome of Cas9 editing. Whether a single targeted DSB efficiently triggers local SCE, how SCE frequency scales when hundreds of sites are cut simultaneously, and whether Cas9-induced SCEs are uniformly error-free remain open questions.

Template-strand sequencing methods, including Strand-seq and its scalable derivatives OP-Strand-seq and sci-L3-Strand-seq, overcome the fundamental limitation that SCE between identical sisters leaves no sequence change (Chovanec et al., 2026; Falconer et al., 2012; Hanlon et al., 2022). By selectively degrading BrdU-labeled nascent DNA and sequencing only parental template strands, these methods detect strand switches genome-wide in single cells, providing a direct readout of non-local repair events independent of any sequence alteration. Furthermore, the high throughput of sci-L3-Strand-seq enables recovery of reciprocal daughter-cell pairs (RDCPs), in which both products of a single mitotic division are captured, allowing direct assessment of whether SCE events are reciprocal and copy-neutral or accompanied by structural alterations.

Here, we use sci-L3-Strand-seq to map Cas9-induced SCE at single-cell resolution across two experimental regimes: a single unique-locus cut and simultaneous cleavage at 237 repetitive targets. We show that a single Cas9 DSB at a unique site is a potent local trigger of SCE. At repetitive loci, on-target SCE enrichment is modest in bulk but becomes significant in a subset of cells with elevated recombination burden. RDCP analysis reveals that Cas9-induced SCEs are not uniformly benign: structural alterations at exchange junctions indicate that repair of two-ended, isochromatid Cas9 DSBs frequently deviates from error-free HR, in contrast to the clean reciprocal outcomes observed for spontaneous, replication-born SCEs in a companion study (Chovanec et al., 2026, submitted). Wild-type Cas9 and nickase variants (D10A, H840A) are compared, and the consequences of disrupting DNA repair genes at the cut site are assessed. Together, these results provide the first genome-wide, single-cell map of Cas9-induced SCE and reveal that the isochromatid geometry of Cas9 breaks creates a distinct recombination landscape with both error-free and mutagenic outcomes.

Results

Cas9 duplex cleavage, but not nicking, can induce durable cell-cycle arrest at single and repetitive targets

To define how Cas9-induced lesions impact cell-cycle progression and SCE, we used recombinant Cas9 ribonucleoprotein (RNP) complexes (IDT), a widely adopted delivery format for genome editing that enables efficient generation of monoclonal knockouts as well as high-efficiency (>90%) polyclonal knockouts for bulk perturbation experiments. We compared four single-cutting sgRNA, each targeting a unique genomic locus, with a multi-target sgRNA recognizing 237 repetitive sites. Introduction of duplex-cleaving Cas9 at either a single site or across repetitive targets led to a pronounced cell-cycle arrest within 24 hours of delivery, indicating that even a single DSB is sufficient to trigger a robust checkpoint response, irrespective of whether the targeted locus resides in an essential gene (LIG3) or a non-essential gene (LIG4). Notably, this arrest persisted after Cas9 removal only in cells targeted with the repetitive sgRNA, consistent with sustained or cumulative DNA damage at multiple genomic sites (Fig.1A-B, Fig.S1A).

Experimental design, cell cycle perturbation and genome-wide pileup of both spontaneous and Cas9-induced on-target SCE events.

A. Experimental design. BrdU was incorporated to enable SCE mapping by sci-L3-Strand-seq, either during the same 24-hour window as Cas9 RNP treatment or during a 24-hour period following Cas9 RNP washout. Wild-type Cas9 was used unless otherwise indicated for nickase variants. B. FACS analysis of cell-cycle progression following Cas9 RNP treatment. “24hrs” denotes concurrent Cas9 RNP treatment and BrdU labeling, whereas “24 ws BrdU” indicates BrdU labeling after Cas9 RNP washout. Each of the four single-cut sgRNAs targets a distinct gene, as indicated. C. Genome-wide pileup plots for both spontaneous and Cas9-induced on-target SCE events. For sgRNAs targeting XRCC1 and XRCC4, SCEs from both the first division during RNP treatment and the post-washout division are shown. For LIG3 and LIG4, SCEs from the first division are shown. Additional pileups, including comparisons between breakpointR-based and rescued SCE calls, are provided in Fig.S1B. D. Quantification of on-target SCE events for the four single-site sgRNAs in the first division (24 hrs) and in the subsequent division following RNP washout (24 ws).

Cas9 nickases, in which one of the two nuclease domains is inactivated (D10A or H840A), were developed to reduce off-target mutagenesis by converting DSBs into a single-strand nicks (Mali et al., 2013; Ran et al., 2013). Although paired nickases with offset guide RNAs can achieve on-target editing with improved specificity, individual nicks are not inherently inert, as they can be converted into DSBs by replication fork run-off or coincident nicking of the opposite strand. Given that only duplex cleavage at repetitive sequences induced a persistent cell cycle arrest after washout, we tested whether nickase-induced lesions at the same sites had a similar effect. As shown in Fig.1B, neither D10A nor H840A nickase variants measurably perturbed cell-cycle progression following washout. These findings indicate that duplex DNA cleavage, particularly when distributed across repetitive loci, imposes a durable cell-cycle barrier, whereas single-strand nicking is largely tolerated under these conditions.

A single Cas9 DSB induces potent local SCE

We next examined the genomic distribution of Cas9-induced and background spontaneous SCEs using genome-wide pileup analysis (Fig.1C, Fig.S1B). Across all four single-targeting sgRNAs, SCEs were strongly enriched at the programmed cut sites, demonstrating robust local induction of strand exchange. This enrichment was most pronounced in the first cell division following RNP delivery and was reduced 24 hours after washout. To account for low-coverage cells in which discrete breakpoints could not be confidently called, we performed a complementary analysis by quantifying strand-specific read imbalance within 10 Mb flanking the cut site, which similarly revealed clear evidence of SCE at the targeted loci. As summarized in Fig.1D (and Fig.S1C), the on-target SCE rate reached up to 41% in the first division and 17% after washout. Overall, the on-target rate declined by approximately half following washout, with the largest reduction of 74%, indicating that Cas9-induced SCE is most efficiently captured immediately after cleavage and diminishes in subsequent cell cycles. Notably, while Cas9 RNP delivery achieves high bulk editing efficiencies (∼90%), this likely reflects multiple rounds of cleavage and repair at the same locus; consequently, the observed SCE frequencies should not be interpreted as a proxy for the final error-free HR or for knockout efficiency. Rather, the SCE rate reported here reflects the fraction of Cas9-induced breaks repaired through inter-sister exchange rather than cis repair on the same chromatid.

Disruption of DNA repair genes at the cut site did not measurably alter SCE frequency per cell

Given the delayed functional loss following Cas9-mediated gene disruption, we do not expect spontaneous SCE to increase as a result of sgRNA-directed cutting within DNA repair genes over the timeframe of these experiments. Even biallelic frameshift mutations do not immediately eliminate the pre-existing protein pool, which must be diluted through turnover and cell division before a functional deficit can emerge. Consistent with this expectation, genome-wide SCE counts per cell were indistinguishable across all four gene-targeting conditions (LIG3, LIG4, XRCC1, XRCC4) at the 24-hour simultaneous labeling timepoint with all showing a median of 3 SCEs per cell, comparable to the wild-type with Cas9 cuts (Chovanec et al., 2026) and Lig4 washout controls (pairwise Wilcoxon p > 0.09 for all 24-hour comparisons among single-cut repair gene-target conditions, Fig.S1C). This contrasts with stable XRCC1 knockout, which increases spontaneous SCE approximately five-fold in our companion study (Chovanec et al., 2026, submitted).

The LIG3 locus provides an informative example. As an essential gene, complete loss of LIG3 is expected to impair cell viability. In the 24-hour simultaneous labeling condition, >91% of recovered cells were in their first division following Cas9 delivery, and we observed no reduction in nuclei recovery, indicating minimal immediate selection. In contrast, in the washout condition, where labeling occurred in the second cycle, nuclei yield was markedly reduced, although the fraction of surviving cells progressing into the subsequent G1 was high. This pattern is consistent with selective loss of cells carrying biallelic loss-of-function alleles in the second cell cycle. The recovered cells, including those with on-target SCE at the LIG3 locus, likely retained functional LIG3, either through error-free repair (including SCE itself, which preserves sequence) or in-frame edits.

More broadly, although Cas9 RNP delivery achieves high bulk editing efficiencies by amplicon sequencing, this does not equate to functional equivalence with siRNA-mediated knockdown in single-cell assays. A population in which ∼90% of cells carry biallelic knockout is not equivalent to one in which all cells experience partial depletion. For essential genes in particular, single-cell readouts are inherently biased toward the surviving, functionally intact fraction, potentially obscuring the phenotype of interest.

On-target SCE enrichment at repetitive loci is mild but concentrated in high-SCE cells

We then asked whether SCE is enriched at the 237 repetitive target sites following Cas9 cleavage. To test this, we compared the number of SCE breakpoints falling within regions of interest (ROIs) flanking programmed cut sites against the number expected from size-matched random genomic intervals, using a Fisher’s exact test (Tab.1). Across all cells, duplex Cas9 cleavage at 237 repetitive targets produced modest on-target SCE enrichment. The strongest signal came from the post-washout BrdU-labeled condition (“237cuts_24ws_BrdU”), which yielded a 1.20-fold enrichment (p = 4.89 × 10E-55, n = 2,547 cells), with an average of 6.13 on-target SCE breakpoints per cell. Adding Cas9 RNP and BrdU labeling simultaneously (“237cuts_24hrs”) showed a similar fold enrichment of 1.10 (p = 0.02, n = 214), with 5.72 on-target SCEs per cell. By contrast, the earlier WT237 experiment showed no significant enrichment (1.11-fold, p = 0.19, n = 74), likely reflecting the smaller sample size. The D10A nickase produced a marginally significant enrichment (1.09-fold, p = 0.01, n = 330), while the H840A nickase showed no enrichment (1.02-fold, p = 0.37, n = 757). These results indicate that duplex cleavage at repetitive loci induces detectable but mild on-target SCE enrichment in the bulk population, while single-strand nicking produces little or no measurable effect.

A stronger pattern emerged when analysis was restricted to cells with significantly elevated SCE count (>=9 SCE per cell). This threshold represents a 2-3 fold increase above the median of 3-4 SCE in non-cut wild-type cells (Chovanec et al., 2026), and approximately two-fold above the median for the respective Cas9-treated populations, in which the median SCE per cell does not increase beyond 4 despite the introduction of up to 237 DSBs. In the post-washout BrdU condition, this high-SCE subset (n = 261 cells) showed 1.39-fold enrichment (p = 1.29 × 10E-41), with an average of 14.61 on-target SCEs per cell, compared to 6.13 in the full population. The simultaneous 24-hour condition, though limited to 4 cells, showed a comparable 1.39-fold enrichment with 11.50 on-target SCEs per cell. That the bulk median SCE remains unchanged despite 237 concurrent DSBs, while a minority of cells shows both elevated total SCE and concentrated on-target enrichment, suggests that most cells resolve Cas9-induced breaks through non-exchange, cis repair, and that inter-sister exchange is conditionally engaged in a SCE-permissive subpopulation. The WT237 (n = 8) and nickase conditions showed no significant enrichment in this subset, consistent with either insufficient statistical power or a genuine absence of on-target SCE in high-SCE cells following nicking. Relaxing the threshold to >=8 SCE per cell yielded qualitatively similar trends, particularly for the 24-hour condition (1.33-fold, p = 0.06, n = 10) and the post-washout condition (1.35-fold, p = 7.92 × 10E-45, n = 395).

Together, these data indicate that Cas9-induced SCE at repetitive targets is not uniformly distributed across the cell population. A subset of cells with elevated SCE burden shows both more on-target breaks (∼2-fold) and a higher fold enrichment of SCE at cut sites relative to random loci (1.39 vs 1.10-1.20 in the bulk), indicating that these cells not only sustain more Cas9 cleavage events but also resolve a greater fraction of them through inter-sister exchange. Whether this reflects a cell-cycle state more permissive to both cutting and recombination, or stochastic variation in RNP uptake coupled with a recombination-prone chromatin environment, remains to be determined.

RDCP analysis reveals large-scale SVs on chromosomes with induced SCE, as well as structural alterations at Cas9-induced SCE junctions

Reciprocal daughter cell pair (RDCP) analysis enables direct recovery of all four chromatids following replication, providing a stringent readout of SCE outcomes and associated structural changes. Leveraging the throughput of sci-L3-Strand-seq, we previously showed that spontaneous SCEs in XRCC1−/− cells are uniformly reciprocal and copy-neutral. This framework allows us to test whether Cas9 RNP-induced lesions similarly resolve without gross rearrangement. Given the limited sampling achievable with single cut sites, we focused RDCP analysis on cells targeted with the 237-site sgRNA. We identified one RDCP under simultaneous labeling conditions, containing two reciprocal SCE events (Fig.2A). Following washout, we recovered 11 RDCPs, of which four exhibited associated structural variants (SVs) (Fig.2B). Across these four pairs, 14 of 26 events were accompanied by SVs, often arising across two consecutive cell divisions, whereas 12 events were reciprocal and copy-neutral. The remaining seven RDCPs comprised a total of 24 exclusively reciprocal, copy-neutral SCEs (Fig.S2). Reconstruction of repair outcomes is detailed in Tab.S1. Our interpretation is further validated by haplotype-aware analysis in Fig.S3. Overall, 14 of 50 SCE events were coupled with SVs, in contrast to 87 of 87 spontaneous SCE events in XRCC1−/− cells, which were uniformly reciprocal and copy-neutral in our companion study (Chovanec et al., 2026, submitted). Additionally, we observed structural alterations at Cas9-induced SCE junctions consistent with the resolution of under-replicated regions (URR) or replication terminal zones as SCE (Fig.2B, Pair 4, Fig.S3D). The mechanistic implications of these observations are explored further in the Discussion.

RDCP analysis with sgRNA targeting 237 genomic sites.

A. A single RDCP identified from 214 single cells following Cas9 RNP treatment targeting 237 sites with concurrent BrdU labeling. B. Four RDCPs exhibiting large-scale SVs recovered from cells labeled with BrdU after Cas9 RNP washout, corresponding to the second division.

Discussion

This study provides the first genome-wide, single-cell map of SCE induced by CRISPR/Cas9. Three principal findings emerge. First, a single Cas9 DSB at a unique genomic locus is a potent trigger of on-target SCE, with 17-41% of cells undergoing inter-sister strand switching within the same division as cleavage and BrdU labeling. When Cas9 is washed out and cells are labeled in the subsequent cell cycle, the on-target SCE rate drops by 50-74% to a range of 4-17%, confirming that the exchange signal reflects acute repair of the programmed, on-target DSB rather than a heritable alteration of SCE propensity. These strand switches represent non-local interactions between two sister chromatids that are invisible to amplicon sequencing and whole-genome sequencing, yet occur at frequencies comparable to those of the indel outcomes conventionally used to measure editing efficiency. Second, directing Cas9 to 237 repetitive targets produces only mild on-target SCE enrichment in the bulk population, but a subset of high-SCE cells shows more significant enrichment, consistent with a stochastic model in which a fraction of cells sustains a disproportionate share of recombinogenic cleavage events. Even in these high-SCE cells, the number of on-target SCEs per cell remains below 15, far fewer than the 237 available cut sites, indicating that most breaks never formed and/or were resolved by pathways that did not produce SCE. Third, Cas9 nickases (D10A and H840A) neither stall the cell cycle nor produce significant on-target SCE enrichment, consistent with the expectation that single-strand nicks are less recombinogenic when delivered to a single genomic site without a paired offset nick.

Among the most informative outcomes in this dataset is a subset of Cas9-induced SCE events that display a distinctive structural signature in RDCP analysis: one daughter cell carries a strand switch with a WWC or WCC template-strand pattern at the SCE junction, while the paired daughter cell carries a deletion at the matching SCE junction. This “WWC-or-WCC/deletion pair” signature is inconsistent with classical homologous recombination, in which both daughters inherit reciprocal, copy-neutral strand switches. Instead, it resembles the products predicted by the URR resolution model, in which TRAIP-dependent unloading of the CMG helicase at stalled or converging replication forks exposes the template strands to nuclease cleavage, generating one-ended DSBs megabases apart, that are then joined by DNA polymerase theta to produce a sister chromatid with a deletion and an accompanying strand switch (Deng et al., 2019; Sonneville et al., 2019). Independently, van Vugt and colleagues demonstrated that SCEs can form in BRCA1/2-null cell lines, and are enriched at common fragile sites, providing cytogenetic evidence for an HR-independent, URR-coupled SCE pathway (Heijink et al., 2022). Because this pathway does not require end resection or strand invasion, it can operate in BRCA1/2-deficient cancer cells.

Our observation of the “WWC-or-WCC/deletion pair” signature in wild-type cells provides, to our knowledge, the first genetic evidence for this mechanism at single-cell genomic resolution, and a model for its formation is illustrated in Fig.3. Several mechanistic questions follow directly. Does TRAIP-mediated CMG unloading license structure-specific nucleases to process the exposed fork? Which resolvases (MUS81-EME1, SLX1-SLX4, GEN1) execute the cleavage? Note that this specific example actually represents lagging strand cleavage and is possible, although unlikely, due to Cas9 cleavage, RDCP analysis can resolve whether the leading or lagging strands are cleaved, which provides an unique opportunity if coupled with resolvase perturbations. Finally, we observe one such signature among approximately 36 otherwise normal SCEs in wild-type cells; determining the rate of this signature in BRCA1/2-null and other HR-perturbed conditions in spontaneous, rather than Cas9-induced SCE, will be essential for understanding the contribution of URR-type resolution to genome instability in cancer.

Haplotype-aware analysis of observed RDCP (Pair 4, chr1) provides direct genetic evidence of predicted RDCP signature of SCE mediated by URR or replication termination zone (green shaded area), although the cut has to happen on the lagging strand.

Oval depicts centromere. D1: daughter cell 1. D1 has a deletion region at the SCE junction. The corresponding region in D2 has a WC region within the haplotype with the SCE (WWC if depicted together with the other haplotype) as both strands are segregated without replication. Details in text.

Limitation

A limitation of this study is that all experiments were performed in wild-type cells. The genetic regulation of Cas9-induced SCE by the NHEJ, microhomology-mediated end joining (MMEJ), and HR pathways remains to be dissected. In our companion paper, we found minimal role of NHEJ in regulating spontaneous SCE. However, it is unknown whether loss of NHEJ would redirect a greater fraction of Cas9 DSBs toward SCE, particularly given that Cas9 DSBs are two-ended, or whether loss of NHEJ abolishes apparent SCEs formed by dNHEJ across sister chromatids. Similarly, the role of MMEJ in competing with or contributing to SCE outcomes at isochromatid breaks and/or at URRs has not been tested. Systematic genetic perturbation in future experiments will be necessary to delineate the pathway hierarchy governing SCE at Cas9-induced isochromatid DSBs.

Conclusion

In summary, this work establishes that Cas9-induced DSBs trigger SCE at frequencies that are high for single-site cuts and detectable, though modest, for multi-target sites. Beyond quantifying on-target SCE rates, RDCP analysis reveals a “WWC-or-WCC/deletion pair” signature that cannot arise from classical homologous recombination and instead implicates a TRAIP/URR-type resolution pathway operating independently of BRCA1/2. This signature, observed here at single-cell genomic resolution in wild-type cells, provides a molecular framework for the HR-independent SCEs previously inferred from cytogenetic studies in BRCA-deficient backgrounds. More broadly, the isochromatid geometry of Cas9 breaks, in which both sisters are cleaved at the same locus, creates a recombination substrate structurally distinct from the single-ended, replication-born lesions that generate spontaneous SCE (Chovanec et al., 2026, manuscript submitted), and this distinction is reflected in a measurable fraction of mutagenic outcomes. These findings underscore that single-cell, strand-level resolution is necessary for a complete accounting of the repair consequences of programmable nucleases.

Methods

Cell culture

BJ-5ta cells (CRL-4001, ATCC) were cultured in DMEM/F12 medium supplemented with 10% fetal bovine serum (FBS) and 1× penicillin-streptomycin at 37 °C in 5% CO2.

Cas9 RNP delivery

Cas9 ribonucleoprotein (RNP) complexes were assembled using Alt-R S. pyogenes Cas9 Nuclease V3 (1081058; IDT) or corresponding nickase variants (Alt-R Cas9 D10A Nickase V3; Alt-R Cas9 H840A Nickase V3) together with Alt-R tracrRNA (1072532; IDT) and Alt-R crRNA.

For single-cut experiments, crRNAs targeted coding exons of four DNA repair genes (LIG4: 5′-GCTTATACGGATGATCATAA-3′; XRCC4: 5′-ATGGTCATTCAGCATGGACT-3′; XRCC1: 5′-TGCAGGACACGACATGG-3′; LIG3: 5′-CTGTTAGGTACACATCACCG-3′). For the 237-cut condition, a crRNA targeting a repetitive element with 237 predicted genomic sites was used (5′-TCAGCACTTTGGGAGACCA-3′). CrRNA and tracrRNA were diluted to 1 μM in duplex buffer (11010301; IDT), annealed at 95 °C for 5 min, and cooled to room temperature for 30 min to form gRNA. Cas9 (or nickase) was diluted to 1 μM in OptiMEM, combined with gRNA and Cas9 PLUS reagent (CRISPRMAX kit, CMAX00003; Invitrogen), and incubated at room temperature for 5 min to assemble RNPs. Reverse transfection was performed with cells at 2×105 cells/mL.

BrdU labeling conditions

Two labeling schemes were used: in the simultaneous condition (“24 hrs”), BrdU (40 μM final concentration) was added at the time of Cas9 RNP delivery, and cells were cultured for 24 hours prior to fixation, such that Cas9 cleavage and BrdU incorporation occurred within the same cell division. In the washout condition (“24 ws”), Cas9 RNP was delivered without BrdU for 24 hours, after which cells were washed and replated in fresh medium containing 40 μM BrdU for an additional 40 hours before fixation, labeling the division following Cas9 activity.

Sci-L3-Strand-seq library generation

Sci-L3-Strand-seq was performed as previously described (Chovanec and Yin, 2025; Yin et al., 2019). Briefly, cells were trypsinized and fixed in 1.5% formaldehyde in PBS at 1×106 cells/mL for 10 min at room temperature, followed by quenching with glycine. Cells were snap-frozen in nuclei freezing buffer (NFB: 50 mM Tris pH 8.0, 25% glycerol, 5 mM Mg(OAc)2, 0.1 mM EDTA, 5 mM DTT, 1× protease inhibitor).

Tagmentation was performed using Tn5 transposase with 24 first-round barcodes as described (Yin et al., 2019). After tagmentation (55 °C, 15 min), reactions were quenched with lysis buffer (LBT) and nuclei were pooled. Ligation with 72 second-round barcodes was followed by staining with Hoechst-33258 (10 ng/μL) and FACS sorting (200-300 nuclei per well into 96-well plates). Plates were stored at -80 °C.

Following gap extension, UV treatment (270 mJ/cm2, 365 nm), USER digestion, and UGI treatment, T7 in vitro transcription was performed (14-16 hours, 37 °C). RNA purification, reverse transcription, second-strand synthesis (third barcode incorporation), and Illumina library preparation were carried out using standard kits (NEB).

SCE analysis

Raw FASTQ files were processed using the sciL3Pipe Snakemake pipeline (v0.2.1), including barcode demultiplexing, alignment, and single-cell BAM generation. Breakpoint identification and filtering were performed using breakpointR (v1.10.0) and sciStrandR (v0.2.3). Cells passing quality filters (background estimate > 0 and ≤ 0.08; strand-neutral < 0.75; median reads per Mb > 2) were retained.

On-target SCE identification

SCEs were classified as on-target by two criteria. First, a breakpointR-called breakpoint within +/-3 Mb of the cut site was counted as a direct call. Second, a rescue procedure examined strand-state in two 10 Mb bins flanking the cut site (each offset 3 Mb from the gene boundary) and called a rescued SCE if the flanking bins showed a strand-state transition (e.g., WW-to-CC, WW-to-WC, CC-to-WC etc.). On-target strand switches were defined as the union of called and rescued events.

On-target SCE enrichment analysis (237-cut condition)

To test enrichment, observed SCEs overlapping cut-site windows were compared to those overlapping size-matched random genomic intervals using a one-sided’s exact test. Analyses were performed across all cells and repeated for high-SCE cells (cells with >=8 or >=9 post-filter SCEs).

Reciprocal daughter cell identification

The genome was partitioned into 1 Mb bins across autosomes and chromosome X. For each cell, W and C reads were counted per bin, and a strand ratio score was calculated as (W - C)/(W + C), yielding +1 (WW) or -1 (CC). Bins with fewer than four reads were set to 0. For each cell pair, a “sisterness” score was computed by averaging |scorei + scorej| across informative bins (|score| > 0.5 in at least one cell). Scores approach 0 for reciprocal strand inheritance and 2 for identical states. The top 20 candidate pairs with the lowest scores were selected and validated by visual inspection of Strand-seq ideograms.

eLife Assessment

Using sci-L3-Strand-seq, this study shows genome-wide, single-cell evidence that DSBs triggered by CRISPR/Cas9 are frequently resolved through sister chromatid exchange (SCE), which is undetectable with conventional whole-genome sequencing approaches. This important work provides helpful metrics quantifying the occurrence of SCE events at targeted and repetitive genomic loci that are of particular interest to those working in gene editing, DNA repair, genome instability, or repetitive genomic loci. However, the evidence supporting the proposed involvement of under-replicated region/replication-termination-zone resolution and TRAIP/URR-like pathways is currently incomplete and could be strengthened with an increased number of reciprocal daughter-cell pairs and by genetic or molecular perturbation, or alternatively, this can be addressed by changing the discussion.

Reviewer #1 (Public review):

Summary:

This manuscript uses sci-L3-Strand-seq to map sister chromatid exchange events following CRISPR/Cas9-induced DNA damage. Because exchanges between identical sister chromatids are largely invisible to conventional sequencing, the study addresses an important blind spot in the assessment of genome editing outcomes. The authors compare single-locus Cas9 cleavage, simultaneous targeting of 237 repetitive genomic sites, and Cas9 nickase variants. They further use reciprocal daughter-cell pair analysis to ask whether Cas9-associated SCEs are copy-neutral or linked to larger structural alterations. Overall, this is a valuable study that introduces an important additional layer to the analysis of CRISPR/Cas9 repair outcomes. The central finding that Cas9-induced DSBs can trigger frequent local SCE is well supported and likely to be of broad interest. The evidence for structural complexity associated with some induced SCEs is intriguing, but the mechanistic interpretation should either be tested directly or presented more cautiously.

Strengths:

The major strength of the manuscript is the application of a strand-resolved, single-cell method to a question that is difficult to address with standard genome sequencing. The evidence that a single Cas9-induced DSB can trigger strong local SCE is compelling in concept and supported by multiple guide RNAs targeting distinct loci. The reported on-target SCE frequencies, reaching up to 41%, suggest that inter-sister exchange is a substantial and underappreciated outcome of Cas9 cleavage.

Of particular interest is the comparison between single-site and multi-site targeting. The finding that 237 programmed Cas9 targets produce only mild bulk enrichment of on-target SCE but stronger enrichment in a subpopulation of cells with elevated SCE burden is interesting and may have wider biological implications, particularly if the findings extend beyond Cas9-induced SCE to spontaneous SCEs. Given that potential, the current manuscript would benefit greatly from any experiments characterizing this sub-population: are these cells in a particular cell cycle state, experiencing changes in gene expression, or do they have other unique biological properties?

The reciprocal daughter-cell pair analysis is another notable feature of the study. The observation that some Cas9-associated SCEs are accompanied by structural alterations could challenge the assumption that SCE after a programmed break reflects error-free homologous recombination.

Weaknesses:

The number of informative RDCPs is limited, and the mechanistic interpretation of the "WWC-or-WCC/deletion" signature is more suggestive than definitive. In particular, the manuscript invokes (even though only in the Discussion section) URR or replication-termination-zone resolution and discusses TRAIP-dependent CMG unloading, nuclease cleavage, and polymerase theta-mediated joining, but these pathway components are not directly tested herein. A more conservative conclusion that some Cas9-associated SCEs coincide with structural alterations is more appropriate, particularly in the Discussion and Conclusion. For example, the statement that this work provides "direct genetic evidence" for a URR-type mechanism is overstated unless supported by additional experiments or a more extensive analysis of alternative models. Similarly, while the authors explain the limitations of acute Cas9 disruption of LIG3, LIG4, XRCC1, and XRCC4, the manuscript should clarify what biological questions this experiment can and cannot answer.

Reviewer #2 (Public review):

Summary:

In this short paper, a clever single-cell Strand-seq method was used to study the number and location of sister chromatid exchange events (SCEs) in cells after CRISPR/Cas9-induced DNA double-strand breaks (DSBs). Unique as well as multiple genomic loci were targeted. Cas9-induced cuts at unique genomic locations led to statistical enrichment of SCEs at the target site, whereas Cas9 targeted at repetitive targets revealed only mild enrichment of on-target SCEs unless analysis was restricted to a subset of cells with >8 SCEs per cell. Interestingly, reciprocal daughter-cell pair analysis revealed large-scale structural alterations on some chromosomes. Whereas disruption of DNA repair genes, including LIG3, LIG4, XRCC1, and XRCC4, did not measurably alter SCE frequency per cell within 24 hrs, consistent with delayed functional loss following editing and selection against essential genes. Together, these findings demonstrate that Cas9-induced DSBs are potent local triggers of SCE at unique loci and can be associated with structural alterations, highlighting the influence of lesion type and genomic context on recombination outcomes during genome editing.

Strengths:

The data in this paper represent a very rich resource of how parental DNA template strands are distributed in paired daughter cells after various treatments. Abnormalities observed in only one of such paired daughter cells provide a novel and exciting approach to study mechanisms of DNA instability and DNA repair at a genome-wide level in general and following Cas9-induced DSB in particular.

Weaknesses:

The effect of Cas9-induced DSBs in the cells that are used will depend on the cell cycle stage of the cells that are targeted, as well as the number of times cuts are made. The latter could happen before, during, and after DNA repair reactions on one or both alleles in a diploid cell. As a result, it is very difficult to extrapolate the mechanisms of DNA instability and DNA repair from the observed genomic rearrangements. Novel approaches are needed to limit the number and timing of Cas9-induced breaks to overcome some of these limitations. The language and logic in the paper can be improved, and some of the claims seem incorrect. For example, the abstract reads "A single Cas9 cut at a unique genomic locus led to strong local enrichment of SCE at the break site, reaching up to 41% in the same cell cycle and 17% in the subsequent division, indicating that DSB repair frequently engages non-local inter-sister repair." The evidence that only a single Cas9 cut was made is lacking (see my earlier comment); it is not clear how local enrichment or non-local inter-sister repair are defined.

Reviewer #3 (Public review):

Summary:

Chovanec and Yin used their newly developed sci-L3-Strand-seq powerful method to characterize SCE after Cas9 cleavage in a human cell line, using either a single target site or an element repeated 237 times in the genome. SCE are often neglected in DNA repair analyses since they are “genetically silent”. Interestingly, the authors found enrichment of SCE at unique Cas9 sites, but only a modest enrichment of SCE when Cas9 targets 237 sites in the genome. The genetic control of SCE formation at Cas9 sites is not deliberately addressed in this paper. However, the authors found that targeted SCE seem to be enriched in a subpopulation of cells, particularly “permissive” for SCE, but the determinants of such a population are unknown. Finally, the power of the sci-L3-Strand-seq allowed the authors to characterize a specific type of SCE based on the analysis of reciprocal daughter-cell pairs' genomes that is associated with a specific type of chromosomal rearrangement compatible with the ones observed in HR defective BRCA1/2 deficient cells.

Strengths:

This is an interesting paper that molecularly explores sister chromatid exchanges, which represent an important challenge in molecular biology since they are genetically silent.

Weaknesses:

A complexity of the current paper is that it heavily relies on a recently published paper (Chovanec et al 2026, NAR) describing the powerful but complex technique sci-L3-Strand-seq. Knowledge of this paper is a prerequisite to understanding the current manuscript because no reminder is provided. In addition, the current manuscript presents the use of the sci-L3-Strand-seq technique in the study of SCE after Cas9-induced DSBs, while a companion study is referred to several times for containing results about SCE in XRCC1 KO. At some point, one questions the relevance of splitting the use of sci-L3-Strand-seq in different papers instead of making a single integrated one.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

Using sci-L3-Strand-seq, this study shows genome-wide, single-cell evidence that DSBs triggered by CRISPR/Cas9 are frequently resolved through sister chromatid exchange (SCE), which is undetectable with conventional whole-genome sequencing approaches. This important work provides helpful metrics quantifying the occurrence of SCE events at targeted and repetitive genomic loci that are of particular interest to those working in gene editing, DNA repair, genome instability, or repetitive genomic loci. However, the evidence supporting the proposed involvement of under-replicated region/replication termination-zone resolution and TRAIP/URR-like pathways is currently incomplete and could be strengthened with an increased number of reciprocal daughter-cell pairs and by genetic or molecular perturbation, or alternatively, this can be addressed by changing the discussion.

We appreciate the editor’s and the reviewers’ recognition of Cas9-induced SCE as an important previously invisible repair outcome and of RDCP analysis as a notable feature of the study. As proposed, we incorporated the two recent Science studies and now tone down the Discussion of our RDCP observations as consistent with, rather than definitive evidence for, the TRAIP-dependent pathway. We also detail how our single-cell genomic observations complement and extend these two studies in terms of the biological significance of the CDK1-TTF2-TRAIP axis.

Specifically, these changes are in:

Discussion. We changed

“Two recent studies revealed how the CDK1-TTF2-TRAIP axis is cell-cycle regulated to trigger mitotic CMG helicase disassembly and fork cleavage: one study showed a two-fold SCE reduction in mouse ES cells (Fujisawa and Labib, 2026), while the other showed that disrupting the TRAIP-TTF2 interaction reduced common fragile site deletions (Can et al., 2026). Our observation of the "WWC-or-WCC/deletion pair" signature in wild-type cells provides, to our knowledge, the first genetic evidence of linking a deletion with SCE and revealing both W and C unreplicated template strands present in the reciprocal daughter cell, consistent with this mechanism at single-cell genomic resolution (illustrated in Fig.3), although this is limited by the observation of only one RDCP.”

While each study highlights the biological significance of the CDK1-TTF2-TRAIP axis individually for SCE and deletions, we show the coupling relationship (and further evidence of unreplicated template strands) in reciprocal daughter cells.

Fig.3. Title and legends. We changed

“Haplotype-aware analysis of observed RDCP (Pair 4, chr1) shows SV patterns at the SCE junctions consistent with the predicted RDCP signature of SCE mediated by URRs or replication termination zones (green shaded area), although the lagging strands, rather than the leading strands, must be resolved to generate these mitotic breaks.”

Reviewer #1 (Public review):

Summary:

This manuscript uses sci-L3-Strand-seq to map sister chromatid exchange events following CRISPR/Cas9-induced DNA damage. Because exchanges between identical sister chromatids are largely invisible to conventional sequencing, the study addresses an important blind spot in the assessment of genome editing outcomes. The authors compare single-locus Cas9 cleavage, simultaneous targeting of 237 repetitive genomic sites, and Cas9 nickase variants. They further use reciprocal daughter-cell pair analysis to ask whether Cas9-associated SCEs are copy-neutral or linked to larger structural alterations. Overall, this is a valuable study that introduces an important additional layer to the analysis of CRISPR/Cas9 repair outcomes. The central finding that Cas9-induced DSBs can trigger frequent local SCE is well supported and likely to be of broad interest. The evidence for structural complexity associated with some induced SCEs is intriguing, but the mechanistic interpretation should either be tested directly or presented more cautiously.

Strengths:

The major strength of the manuscript is the application of a strand-resolved, single-cell method to a question that is difficult to address with standard genome sequencing. The evidence that a single Cas9induced DSB can trigger strong local SCE is compelling in concept and supported by multiple guide RNAs targeting distinct loci. The reported on-target SCE frequencies, reaching up to 41%, suggest that inter-sister exchange is a substantial and underappreciated outcome of Cas9 cleavage.

Of particular interest is the comparison between single-site and multi-site targeting. The finding that 237 programmed Cas9 targets produce only mild bulk enrichment of on-target SCE but stronger enrichment in a subpopulation of cells with elevated SCE burden is interesting and may have wider biological implications, particularly if the findings extend beyond Cas9-induced SCE to spontaneous SCEs. Given that potential, the current manuscript would benefit greatly from any experiments characterizing this sub-population: are these cells in a particular cell cycle state, experiencing changes in gene expression, or do they have other unique biological properties?

The reciprocal daughter-cell pair analysis is another notable feature of the study. The observation that some Cas9-associated SCEs are accompanied by structural alterations could challenge the assumption that SCE after a programmed break reflects error-free homologous recombination.

Thank you very much for this assessment.

Weaknesses:

The number of informative RDCPs is limited, and the mechanistic interpretation of the "WWC-orWCC/deletion" signature is more suggestive than definitive. In particular, the manuscript invokes (even though only in the Discussion section) URR or replication-termination-zone resolution and discusses TRAIP-dependent CMG unloading, nuclease cleavage, and polymerase theta-mediated joining, but these pathway components are not directly tested herein. A more conservative conclusion that some Cas9-associated SCEs coincide with structural alterations is more appropriate, particularly in the Discussion and Conclusion. For example, the statement that this work provides "direct genetic evidence" for a URR-type mechanism is overstated unless supported by additional experiments or a more extensive analysis of alternative models. Similarly, while the authors explain the limitations of acute Cas9 disruption of LIG3, LIG4, XRCC1, and XRCC4, the manuscript should clarify what biological questions this experiment can and cannot answer.

Please see response to the eLife Assessment as this is a common point raised by multiple reviewers.

Additionally, we clarified what the DNA repair gene targeting experiment can and cannot answer (delayed protein loss, essential-gene selection) by adding the following text in the “Disruption of DNA repair genes at the cut site did not measurably alter SCE frequency per cell” section:

“One additional complication is that, although Cas9 RNP achieves >90% knockout efficiency in bulk assays and is therefore used as a substitute for siRNA, sorting BrdU-labeled cells in the subsequent G1 enriches for cells that escaped frameshift editing, particularly for essential genes; thus, 100% knockout in 90% of cells is not equivalent to 90% knockdown in every cell, representing a unique challenge for single-cell assays.”

Reviewer #1 (Recommendations for the authors):

(1) Temper the mechanistic claims about URR/TRAIP-type resolution.

The RDCP data support the conclusion that some Cas9-associated SCEs are accompanied by structural alterations and may arise through non-classical mechanisms. However, claims about TRAIPdependent CMG unloading, URR resolution, or polymerase theta-mediated joining should be framed as a model unless directly tested.

We cited mechanistic dissection of the CDK1-TTF-2-TRAIP axis, which was published since the review of the paper. While each study highlights the biological significance of the CDK1-TTF2-TRAIP axis individually for SCE and deletions, we show the coupling relationship (and further evidence of unreplicated template strands) in reciprocal daughter cells but qualified that this observation is in only one RDCP. We further tempered our claims by changing “direct evidence” to “consistent with” as we did not perturb genes involved in these processes.

(2) Clarify the impact of the small RDCP sample size.

The manuscript would be strengthened by explicitly stating how many total RDCPs were analyzed, how many SCE events were informative, and how much confidence can be placed in the estimated fraction of SCEs associated with SVs. A short table summarizing RDCP counts, SCE counts, copy-neutral events, and SV-associated events would be helpful.

A total of 15 RDCPs were recovered from close to 4,000 single cells analyzed across all conditions. We added Tab.S3 detailing SCEs in all 15 RDCPs in addition to SCE and SV breakdown in Tab.S2 (originally Tab.S1). The new Tab.S3 is cited in the “RDCP analysis reveals large-scale SVs on chromosomes with induced SCE, as well as structural alterations at Cas9-induced SCE junctions” section.

(3) Provide more detail on the "rescued" SCE calls.

Because the central conclusions rely on SCE detection, the criteria for breakpoint R-based calls versus rescued calls should be explained clearly in the main text or methods. It would be useful to know how sensitive the main conclusions are to the inclusion or exclusion of rescued calls. Is this laid out in greater detail in an additional manuscript?

We previously included an “On-target SCE identification” section in the Methods, where we described in detail the rescue of on-target SCEs missed by the initial breakpoint R calls. We also depicted calls and calls+rescues for all the conditions in Fig.S1B.

We agree with the reviewer and now expanded the description of rescued SCE calls in the main text (under the “A single Cas9 DSB induces potent local SCE” section). In brief, 50-92% of SCEs (typically >70%) across the four single-targeting sites were directly called rather than rescued, with the exception of LIG4, where only 30% were direct calls. This is because LIG4 is located only 6 Mb from the telomere and is therefore particularly prone to missed breakpointR calls in low-coverage cells. On-target rescue at individual sites is self-contained in this manuscript because our lab primarily focuses on spontaneous SCE, for which there are no expected SCE sites. However, the rescue methodology was previously implemented in the original development of sci-L3-Strand-seq to identify SCEs at centromeres.

(4) Clarify the biological interpretation of the repetitive-target enrichment.

The high-SCE subset analysis is interesting, but the manuscript should explain whether these cells have evidence of higher RNP uptake, altered cell-cycle state, greater DNA damage, or lower sequencing quality. If these possibilities cannot be distinguished, the text should state this clearly.

We agree with the reviewer. The high SCE subset does not have lower sequencing quality by coverage or background (0.3% coverage for high-SCE vs. 0.28% coverage overall, and 3% background for both high-SCE and overall). However, our current data do not allow us to distinguish among biological explanations for the elevated SCE. The original manuscript acknowledged this limitation (“Whether this reflects a cell-cycle state more permissive to both cutting and recombination, or stochastic variation in RNP uptake coupled with a recombination-prone chromatin environment, remains to be determined.”). To make this limitation more explicit and to address the possibility of sequencing quality raised by the reviewer, we have revised the text as follows: “The high-SCE subset did not show evidence of lower sequencing quality, based on either sequencing coverage (p=0.17) or background SCE levels (p=0.13). However, we cannot distinguish whether the elevated on-target SCE reflects a cell-cycle state more permissive to both cutting and recombination, or stochastic variation in RNP uptake coupled with a recombination-prone chromatin environment.”

This question may be better explored by future co-assays with sci-L3-Strand-seq; currently we cannot enrich for cells with high SCEs to characterize the molecular features of this subset of the cells using other omics approaches.

(5) Reconsider the framing of the DNA repair gene targeting experiment.

The current data do not strongly test whether LIG3, LIG4, XRCC1, or XRCC4 regulate Cas9-induced SCE, because functional protein loss is delayed and essential-gene targeting introduces selection. This section may be better framed as a negative/control observation rather than as a pathway analysis.

We agree and please refer to Public Reviews for a single-cell assay-specific explanation.

(6) Consider including some additional control experiments, for example, Cas9 without sgRNA, nontargeting sgRNA, or mock-transfected cells to make sure that some phenotypes (for example, cell-cycle arrest) directly result from DNA cleavage rather than from the transfection procedure.

We thank the reviewer for this suggestion. We have carefully considered these additional controls but have chosen not to add further experiments. Our existing Cas9 nickase experiments provide a control that directly addresses whether the observed arrest is attributable to DSB formation rather than RNP delivery/transfection. Both the D10A and H840A Cas9 nickases were delivered under the same experimental conditions as wild-type Cas9, but neither produced the cell-cycle arrest observed following DSB induction by wild-type Cas9. Thus, these experiments control for Cas9 RNP delivery while altering the nature of the DNA lesion and support the interpretation that the observed arrest is associated specifically with Cas9-induced DSBs rather than the transfection procedure itself.

(7) Figure 1: the fonts should be increased. The majority of the labels are impossible to read in a printed copy of this manuscript.

We thank the review for pointing this out. We enlarged Fig.1 fonts.

(8) Figure 1C. The pileup plots should be described and interpreted in a clear way. In its present form, it is unclear how the interpretations and conclusions are made.

We added explanation of the pileup analysis immediately following mentioning the Fig.1C pileup: “We next examined … SCEs using genome-wide pileup analysis (Fig.1C, Fig.S1B), in which we plot the total number of SCEs detected across all single cells within each 1 Mb window.”

Reviewer #2 (Public review):

Summary:

In this short paper, a clever single-cell Strand-seq method was used to study the number and location of sister chromatid exchange events (SCEs) in cells after CRISPR/Cas9-induced DNA double-strand breaks (DSBs). Unique as well as multiple genomic loci were targeted. Cas9-induced cuts at unique genomic locations led to statistical enrichment of SCEs at the target site, whereas Cas9 targeted at repetitive targets revealed only mild enrichment of on-target SCEs unless analysis was restricted to a subset of cells with >8 SCEs per cell. Interestingly, reciprocal daughter-cell pair analysis revealed largescale structural alterations on some chromosomes. Whereas disruption of DNA repair genes, including LIG3, LIG4, XRCC1, and XRCC4, did not measurably alter SCE frequency per cell within 24 hrs, consistent with delayed functional loss following editing and selection against essential genes. Together, these findings demonstrate that Cas9-induced DSBs are potent local triggers of SCE at unique loci and can be associated with structural alterations, highlighting the influence of lesion type and genomic context on recombination outcomes during genome editing.

Strengths:

The data in this paper represent a very rich resource of how parental DNA template strands are distributed in paired daughter cells after various treatments. Abnormalities observed in only one of such paired daughter cells provide a novel and exciting approach to study mechanisms of DNA instability and DNA repair at a genome-wide level in general and following Cas9-induced DSB in particular.

Thank you very much for this assessment.

Weaknesses:

The effect of Cas9-induced DSBs in the cells that are used will depend on the cell cycle stage of the cells that are targeted, as well as the number of times cuts are made. The latter could happen before, during, and after DNA repair reactions on one or both alleles in a diploid cell. As a result, it is very difficult to extrapolate the mechanisms of DNA instability and DNA repair from the observed genomic rearrangements. Novel approaches are needed to limit the number and timing of Cas9-induced breaks to overcome some of these limitations. The language and logic in the paper can be improved, and some of the claims seem incorrect. For example, the abstract reads "A single Cas9 cut at a unique genomic locus led to strong local enrichment of SCE at the break site, reaching up to 41% in the same cell cycle and 17% in the subsequent division, indicating that DSB repair frequently engages non-local inter-sister repair." The evidence that only a single Cas9 cut was made is lacking (see my earlier comment); it is not clear how local enrichment or non-local inter-sister repair are defined.

We agree with the limitations that Cas9-induced DSBs can be dependent on the cell cycle stage and the number of times cuts are made. We also agree that novel approaches are needed to limit the number and timing of Cas9-induced breaks to overcome these limitations, perhaps by using vfCas9 but more importantly, if new approaches to turn off Cas9 are developed. We have added a brief discussion on this limitation (in the Limitation section) and the resulting constraints on extrapolating mechanisms of DNA instability and repair from the observed genomic rearrangements. We thank the reviewer for pointing out the distinction between “a single Cas9 cut” vs. "Cas9 targeting of a single genomic locus." We went through the manuscript and revised where cutting only once was implied. We also explicitly acknowledge the possibility of multiple rounds of cutting at the same sites.

We thank the reviewer for pointing out that “non-local repair” is a non-standard term. We use it operationally to distinguish repair confined to the broken chromatid (e.g., fill-in synthesis or end joining in cis) from repair involving exchange between sister chromatids. We have added a schematic (Fig. S1A) illustrating this distinction and cited this figure immediately before where we operationally defined SCE as a “reciprocal strand switch between sister chromatids, without implying a single mechanistic pathway.” This distinction is important because, particularly for two-ended Cas9 DSBs, an SCE-like outcome could potentially arise through either HR-mediated crossover or NHEJ of DNA ends across sister chromatids; the latter may involve different genetic requirements from classical NHEJ at least in end-tethering. We have revised the manuscript to define “non-local repair” explicitly at its first use.

Reviewer #2 (Recommendations for the authors):

References to relevant earlier studies using Strand-seq to study SCEs are missing (PMID: 27185886 and PMID: 29348659).

We thank the reviewer for pointing this out. We added these references in the 3rd paragraph of the Introduction where we briefly review Strand-seq methods.

Reviewer #3 (Public review):

Summary:

Chovanec and Yin used their newly developed sci-L3-Strand-seq powerful method to characterize SCE after Cas9 cleavage in a human cell line, using either a single target site or an element repeated 237 times in the genome. SCE are often neglected in DNA repair analyses since they are « genetically silent ». Interestingly, the authors found enrichment of SCE at unique Cas9 sites, but only a modest enrichment of SCE when Cas9 targets 237 sites in the genome. The genetic control of SCE formation at Cas9 sites is not deliberately addressed in this paper. However, the authors found that targeted SCE seem to be enriched in a subpopulation of cells, particularly « permissive » for SCE, but the determinants of such a population are unknown. Finally, the power of the sci-L3-Strand-seq allowed the authors to characterize a specific type of SCE based on the analysis of reciprocal daughter-cell pairs' genomes that is associated with a specific type of chromosomal rearrangement compatible with the ones observed in HR defective BRCA1/2 deficient cells.

Strengths:

This is an interesting paper that molecularly explores sister chromatid exchanges, which represent an important challenge in molecular biology since they are genetically silent.

Thank you very much for this assessment.

Weaknesses:

A complexity of the current paper is that it heavily relies on a recently published paper (Chovanec et al 2026, NAR) describing the powerful but complex technique sci-L3-Strand-seq. Knowledge of this paper is a prerequisite to understanding the current manuscript because no reminder is provided. In addition, the current manuscript presents the use of the sci-L3-Strand-seq technique in the study of SCE after Cas9-induced DSBs, while a companion study is referred to several times for containing results about SCE in XRCC1 KO. At some point, one questions the relevance of splitting the use of sci-L3-Strandseq in different papers instead of making a single integrated one.

We appreciate this concern. The original sci-L3-Strand-seq study is an extensive methodology paper that establishes and validates various computational framework, whereas the companion study focuses on the genetic regulation of spontaneous SCE. The experimental designs and biological questions of the companion study and the present work are therefore distinct, although we draw on selected results from the companion study where they provide useful comparisons and contrasts between spontaneous and Cas9-induced SCE. The present study addresses a distinct biological question, the response to Cas9-induced DSBs, and we therefore believe that combining these studies would make the resulting manuscript unnecessarily broad and obscure their different biological questions.

We nevertheless agree that the present manuscript should be understandable without requiring detailed knowledge of either paper. We have therefore added a brief description of the sci-L3-Strandseq approach (3rd paragraph of Introduction, Fig.S1A legends, and Fig.1B legends) and clarified the relevant methodological concepts where they are first introduced. We hope to improve the self-contained nature of the manuscript so that readers need not consult the earlier NAR papers, and the companion preprint to understand the key results.

Reviewer #3 (Recommendations for the authors):

(1) Abstract

"Identical sisters ": redundant

"non-local" inter-sister repair: the meaning is not clear. Do the authors refer only to "inter-sister" and therefore "non-local" is redundant, or do they imply something specific by "non-local", in which case it needs to be clarified?

"237 repetitive targets": at least a slight description of this target is needed. Is it a "random" repeat, a satellite sequence, a sequence related to a transposable element ?...

We agree with the reviewer that “identical sisters” is technically redundant. However, we have retained “identical” here to emphasize the distinction between sister chromatids vs. homolog, as SCE is sometimes misconstrued as exchange between homologs and as potentially causing loss of heterozygosity. We prefer the slight redundancy here for conceptual clarity.

We thank the reviewer for pointing out that the meaning of “non-local” was unclear. As discussed in our response to Reviewer #2, we use “non-local repair” operationally to distinguish repair confined to the broken chromatid in cis from repair involving exchange between sister chromatids. We have added a schematic (Fig.S1A) illustrating this distinction and explicitly define the term at its first use in the revised manuscript. Please see our response to Reviewer #2 above for the detailed rationale.

We thank the reviewer for asking us to clarify the nature of the 237 repetitive targets. The sgRNA targets an Alu sequence and was selected from a larger screen of >20,000 sgRNAs targeting repetitive sequences occurring at >200 genomic sites. In that screen, cellular toxicity did not simply scale with the number of predicted target sites; we therefore selected this sgRNA because its intermediate phenotype allowed us to introduce a large number of programmed DSBs without either minimal perturbation or excessive loss of cells. Thus, the 237-site guide was not an arbitrarily selected Alu-targeting sgRNA. The full repetitive-element screen is beyond the scope of the present study, but we have clarified in the Abstract that these 237 sites are Alu targets and added a brief description of the guide selection in the Methods.

(2) Introduction:

"non-local outcome / non-local repair processes": The use of "non-local" is not standard and is obscure for the reader. Specify if it has any meaning or remove it.

Please see our response above to both Reviewers #2 and #3 regarding our definition and use of “nonlocal repair”.

The authors mention that replication through a DSB generates four broken ends. However, in case the DSB is reached by one replication fork before the converging one, there are only three broken ends for at least the time required for the converging fork to reach the DSB from the other side. This may influence the repair outcome.

We agree with the reviewer. If one replication fork encounters the DSB before the converging fork, a transient three-ended intermediate can exist before the second fork reaches the break. This temporal asymmetry could influence repair pathway choice, including engagement of HR, end joining, or BIR-like repair. Our assay captures the resulting SCE outcome but cannot distinguish the order in which replication forks encounter the DSB or the repair pathway engaged at these intermediate stages. We have revised the text (2nd paragraph of the Introduction) to clarify that four broken ends represent the eventual configuration after replication through the DSB, rather than necessarily a simultaneous intermediate.

(3) Results

Cell cycle arrest experiment: it seems that a control condition with no Cas9 is missing to conclude better about what looks like a G2-M arrest, but that is not clearly mentioned.

Please see our response to Reviewer #1, Recommendation 6, regarding additional controls for the cell-cycle arrest experiment. Briefly, the D10A and H840A Cas9 nickases were delivered under the same experimental conditions as wild-type Cas9 but did not produce the cell-cycle arrest observed following DSB induction, providing an internal control for RNP delivery/transfection and supporting the association of the arrest with Cas9-induced DSBs. We would also like to clarify that the observed cell-cycle arrest is primarily a G1/S, rather than G2/M, basing on the FACS signal (see revised Fig.1B legend). This is consistent with the strong G1/S checkpoint in mammalian cells and the predominantly G1 cell-cycle distribution of BJ-5ta cells.

Note that the font size in Figure 1 is too small for readability.

We have enlarged the font sizes throughout Figure 1 to improve readability.

Figure 1B, D10A and H840A conditions:

The authors mention that nicks can be converted into DSBs through the passage of the replication fork, but do not see any cell cycle defect in the conditions tested. Is it possible that the absence of effect results from the fact that the analysis is done prior to nicks being converted into DSBs? This remark notably applies to the 237 target sites experiment. It seems that controlling for cell cycle delays for longer times is needed to conclude clearly about this aspect. In case a clear absence of cell cycle delay is observed in the 237 target sites in the Cas9 nicking condition, this would suggest that replication born DSBs behave differently from "classical" two-ended DSBs and do not trigger cell cycle arrest.

We agree that the timing of nick conversion during replication could contribute to the absence of a detectable cell-cycle delay under the conditions examined. However, extending the duration of Cas9 nickase treatment or labeling would not necessarily resolve this question, because persistent Cas9 activity permits repeated rounds of nicking across successive cell cycles, making it difficult to relate a later cell-cycle phenotype to a defined replication-born lesion. More generally, we believe that the relationship between replication-associated nicks, SCE formation, and cell-cycle progression is better addressed in the context of spontaneous SCE, which is the focus of our companion study. The present study is focused on SCE following programmed Cas9-induced DSBs, and analysis of replication-born nick lesions would require precise temporal control (ideally vfCas9 nickases that can be turned off) of individual nicking events relative to replication, for which an appropriate experimental system is not currently available to us.

Figure 1C should mention somewhere the genomic location of the four targets to clearly show that they correspond to the four major SCE peaks. In addition, there is no legend for the vertical pink stripes. Finally, it might be wise to keep the same y-axis scale for better comparisons.

Figure 1 overall: it might be wise to clearly show a no Cas9 condition to clearly set the SCE baseline and show that it is independent of Cas9 induction. As of now, it is not clear whether the non-targeted SCE comes from a specific cleavage of Cas9 or not. Such an aspect could benefit from putting Figure S1C in the main Figure 1. Alternatively, results from Chovanec et al 2026 (NAR) should be better restated because the reader does not necessarily have them in mind.

We thank the reviewer for these suggestions. The expected Cas9 target positions were already indicated by vertical bars in Fig. 1C; however, we agree that this was not sufficiently clear. We have therefore revised the figure legend to explain that the vertical bars indicate the expected Cas9 target positions. The Chovanec et al. (2026, NAR) study focused entirely on spontaneous SCE, which we simultaneously map here as the background signal, rather than the on-target SCEs induced by Cas9. We hope that explicitly identifying the target locations in the revised legend makes this distinction clear and ensures that prior knowledge of the NAR study is not necessary to interpret Fig. 1C.

We have retained the individual y-axis scales because the magnitude of SCE enrichment differs substantially among conditions. Using a common y-axis scale would make several of the on-target SCE peaks difficult to visualize.

The section « Disruption of DNA repair genes at the cut site did not measurably alter SCE frequency per cell » is questionable in the results section for the following reasons:

(i) The DNA repair genes are used here as target sites for Cas9 cleavage, but are not the object of the study, but may be the object of a companion paper. This aspect is slightly misleading.

(ii) As first mentioned in this section, there is evidence strongly suggesting that inactivating DNA repair genes will not affect SCE, and this is what the authors observed.

(iii) As an alternative, one could put the emphasis on the fact that the effect of Cas9-mediated inactivation of DNA repair genes (ie LIG3) starts to be detectable only in the washout condition ie after at least one cell cycle. But in this case, this is addressing the role of DNA repair genes in Cas9-induced SCE, which is not the point of the current paper.

We thank the reviewer for this comment and agree that the original framing of this section could give the impression that these experiments were intended to test the functions of the targeted DNA repair genes in SCE. This was not our intent. Rather, these genes provided defined genomic target sites for Cas9 cleavage, and the primary purpose of the experiment was to characterize SCE associated with Cas9-induced DSBs at these loci.

As discussed in our response to the Editor Assessment, there are important limitations to using these experiments to infer the consequences of loss of the targeted proteins, including the delay between Cas9 cleavage and depletion of pre-existing protein and, and particularly in the single-cell assay, selection for cells that escape disruptive editing at essential genes. We have added text to the Results explicitly describing these limitations.

We therefore agree with the reviewer that the delayed effects observed under the washout condition should not be interpreted here as establishing a role for individual DNA repair genes in Cas9-induced SCE. We have revised the section title to “Cas9 targeting of DNA repair gene loci did not immediately alter overall SCE frequency per cell” to clarify the scope of this experiment and to avoid implying that testing the functions of the targeted DNA repair genes is a major objective of the present study.

The conditions in Table 1 need to be homogenized and better explained:

- 237 sites and 237 cuts are used: homogenize?

We thank the reviewer for spotting this. We revised both to be “237 sites”.

- May explain better the rationale for putting BrdU simultaneously with Cas9 or after 24 h and a wash.

For the single-targeting sites, we observed more SCE when BrdU was added simultaneously with the Cas9 for the same 24 hours, compared to adding BrdU in the subsequent division after a wash. Therefore, for the 237 sites, we analyzed both conditions.

- Typo in the text: 237cuts_24ws_40BrdU instead of 237cuts_24ws_BrdU

We apologize for the lack of clarity in these labels and have substantially revised the Table 1 legend. In brief, the “40” is not a typo. In the 237 sites experiments, wild-type Cas9 considerably prolonged the cell cycle. Therefore, rather than labeling with BrdU for 24 hours as in the other conditions, we extended BrdU labelling to 40 hours in the last two conditions to allow more cells to progress into the subsequent G1 for successful Strand-seq analysis. We clarified that “237 sites 24 + 16hrs BrdU” refers to the condition in which Cas9 RNP and BrdU were added simultaneously. After 24 hours of Cas9 RNP treatment, BrdU labeling was continued for an additional 16 hours (a total of 40 hours of BrdU). The “237sites 24ws40BrdU” condition is the corresponding washout condition, in which Cas9 RNP was removed after 24 hours and cells were then labeled with BrdU for 40 hours post-washout.

- 237 cuts: Are some sites more enriched in SCE than others?

Yes, some sites showed greater SCE enrichment than others. We tested whether this variation correlated with chromatin accessibility but found no significant association. This was not unexpected, as the sgRNAs predominantly target Alu elements.

- Table 1: There is a difference between 237 sites 24ws24BrdU and 237cuts24ws40BrdU, with a significant enrichment of on-target SCE for the latter condition only. Could the increase in SCE rise even more with longer BrdU exposure? In other words, does the low enrichment in SCE at target sites in the 237 sites experiment result from a non-optimal timing for the analysis?

Yes, this is possible. We did not systematically test additional treatment or labeling durations. A 24-hour Cas9 RNP treatment is typically used for Cas9 RNP-mediated knockout experiments, and we therefore initially used this duration to assess gene-editing outcomes. For Strand-seq, BrdU labeling is ideally limited to approximately one cell division. Because BJ-5ta cells have an approximately 24-hour cell cycle, extending BrdU labeling substantially beyond 40 hours could allow some cells to undergo a second round of replication and become double-labelled. We therefore did not extend BrdU labeling beyond 40 hours. Thus, the lower enrichment in the 24ws24BrdU condition may in part reflect the timing of the assay.

- Figure 2 / RDCP analysis:

Interpretation of this figure relies exclusively on the 2026 NAR paper from the authors. This, at least, should be mentioned to help the reader understand it. Once the legend restates, this figure misses clear identification of the SCE and other genomic rearrangements. For readability, maybe the full genome should be kept for the supplementary data, and only the rearranged chromosomes should be kept in the main figure so that the rearrangements are clearly visible and annotated.

We thank the reviewer for this suggestion. To make the Strand-seq plots interpretable without relying on our 2026 NAR paper, we have added an explanation of Strand-seq orientation in the third paragraph of the Introduction and in Fig.S1A. We have revised Fig.2 legends to improve readability. We have retained the whole-genome view because Strand-seq data are conventionally presented in this format and it provides important genome-wide context for interpreting the observed events. The rearranged chromosomes and events were annotated in Fig.S3.

(4) Discussion

- Most DSB never formed or did not produce SCE: how to understand this better? What would be the argument in favor of one or the other possibility?

We agree that these are two possible explanations that cannot be distinguished by the current experiment. The absence of an SCE at a targeted site could reflect either inefficient DSB formation or repair of a DSB through a pathway that does not generate an SCE. Distinguishing these possibilities would require direct measurement of cutting efficiency at individual target sites, which was beyond the scope of this study.

- The conclusion about the effect of the Cas9 nickases needs to be toned down as long as the proper timing for SCE analysis has not been performed (see comment above).

We agree and have toned down this conclusion by specifying that no significant on-target SCE enrichment was detected under the conditions tested and acknowledging that we cannot exclude SCE formation at other time points (Discuss, first paragraph).

- As much as possible, avoid the use of non-conventional acronyms like URR.

We agree and have reduced the use of non-conventional acronyms where possible. We have retained URR (under-replicated region), as the term appears seven times throughout the manuscript, but have ensured that it is clearly defined at first use.

- The discussion about the RDCP analysis in the second paragraph of the discussion should refer to Figure 3.

Thank you for pointing this out. We added this reference to Fig.3

Version 2

Abstract

CRISPR/Cas9-induced DNA double-strand breaks (DSBs) are widely used for genome engineering, yet their capacity to provoke sister chromatid exchange (SCE) and associated genome instability remains incompletely understood, in part because exchanges between identical sister chromatids leave no sequence change and are invisible to conventional whole-genome sequencing. Using sci-L3-Strand-seq, a scalable single-cell template-strand sequencing platform for genome-wide SCE mapping, we quantified strand-switch events following targeted Cas9 cleavage. Cas9 targeting of a unique genomic locus led to strong local enrichment of SCE at the break site, reaching up to 41% in the same cell cycle and 17% in the subsequent division, indicating that DSB repair frequently engages non-local inter-sister repair. Extending the assay to 237 repetitive Alu targets using one Alu-specific sgRNA revealed mild and marginally statistically significant enrichment of on-target SCE across wild-type Cas9 and nickase variants (D10A and H840A). However, restricting analysis to a subset of cells with elevated SCE burden (>8 SCEs per cell) uncovered significant enrichment at programmed cut sites, suggesting that recombination at repetitive loci is conditionally engaged in highly recombinogenic cells. Reciprocal daughter-cell pair analysis further revealed large-scale structural alterations on chromosomes with induced SCEs and at SCE junctions, indicating that Cas9-induced SCEs do not universally reflect error-free homologous recombination. Disruption of DNA repair genes at the cut site, including LIG3, LIG4, XRCC1, and XRCC4, did not measurably alter SCE frequency per cell, consistent with delayed functional loss following editing and, in the case of essential genes such as LIG3, selection against disruptive alleles. Together, these findings demonstrate that Cas9-induced DSBs are potent local triggers of SCE at unique loci, while at repetitive regions SCE is more context-dependent and can be associated with structural alterations, highlighting the influence of lesion type and genomic context on recombination outcomes during genome editing.

Introduction

CRISPR/Cas9 genome editing relies on the induction of site-specific DNA double-strand breaks (DSBs) to trigger endogenous repair pathways (Cong et al., 2013; Ran et al., 2013). The two principal repair outcomes, non-homologous end joining (NHEJ) and homologous recombination (HR), have been extensively characterized at the sequence level in wild-type cells and with DNA repair gene perturbations (Chen et al., 2019; Hussmann et al., 2021; Shen et al., 2018): NHEJ generates small insertions and deletions at the break site, while HR can incorporate an exogenous donor template for precise editing (Richardson et al., 2016). However, this framework captures only the local and mutational fraction of the repair landscape. For non-local mutational outcomes, the structural consequences of Cas9 cutting extend well beyond the small indels detected by standard amplicon sequencing. Large deletions spanning kilobases have been documented at on-target sites (Kosicki et al., 2018), and Cas9-induced DSBs can trigger chromothripsis through micronucleus formation and chromosome bridge breakage (Leibowitz et al., 2021). Chromosome loss and segmental aneuploidies have been observed following Cas9 editing in human embryos, attributed to the isochromatid nature of the break (Zuccaro et al., 2020). At the cytogenetic level, chromosomal instability and rearrangements in edited cancer cell lines are frequently missed by PCR-based genotyping (Regan et al., 2025). These observations establish that Cas9 DSBs can produce a spectrum of structural alterations, yet the relationship between SCE and these larger-scale events has not been examined in the same experimental framework.

In this study, we assay what is likely the most frequent non-local outcome (Figure S1A) induced by Cas9: sister chromatid exchange (SCE). Here, we operationally define SCE as a reciprocal strand switch between sister chromatids, without implying a single mechanistic pathway. SCE can therefore arise from inter-sister homologous recombination (HR), double non-homologous end joining (“dNHEJ”) across two isochromatid DSBs, or other non-local repair processes. This distinction is particularly relevant for Cas9, which typically cleaves both sister chromatids at the same locus and thereby generates the precise isochromatid substrate on which dNHEJ can act. If only one sister is cleaved, the result is a two-ended, post-replicative DSB with an intact sister available as a repair template, an ideal substrate for HR-mediated SCE. More commonly, however, Cas9 can generate isochromatid DSBs through two routes: by cleaving both post-replicative sisters independently, or by cutting the unreplicated template ahead of the fork, such that the break is replicated into DSBs on both daughters. In either case, ultimately four broken ends are produced, although not necessarily simultaneously, and neither sister is intact (Zuccaro et al., 2020). The repair outcome then depends on how the four ends are resolved: cis joining (each sister re-ligated to its own ends) restores the original configuration, while trans joining between sisters (the proximal end of one sister joined to the distal end of the other) produces a strand switch operationally indistinguishable from HR-mediated SCE. Whether Cas9-induced SCE at isochromatid breaks arises through HR, through trans end joining (dNHEJ), or through a mixture of both has not been determined. In all cases, exchanges between identical sisters are genetically silent and invisible to conventional whole-genome sequencing, and have therefore been largely overlooked as an outcome of Cas9 editing. Whether a single targeted DSB efficiently triggers local SCE, how SCE frequency scales when hundreds of sites are cut simultaneously, and whether Cas9-induced SCEs are uniformly error-free remain open questions.

Template-strand sequencing methods, including Strand-seq and its scalable derivatives OP-Strand-seq and sci-L3-Strand-seq, overcome the fundamental limitation that SCE between identical sisters leaves no sequence change (Chovanec et al., 2026b; Falconer et al., 2012; Hanlon et al., 2022; van Wietmarschen et al., 2018; van Wietmarschen and Lansdorp, 2016). By incorporating BrdU into newly synthesized DNA during one cell cycle, selectively degrading the BrdU-labeled nascent strands, and sequencing only parental template strands, these methods map the Watson- and Crick-oriented parental strands inherited by each daughter cell. SCEs are detected as switches in parental strand orientation along a chromosome, providing a direct genome-wide readout of non-local repair events in single cells independent of any sequence alteration (Figure S1A). Furthermore, the high throughput of sci-L3-Strand-seq enables recovery of reciprocal daughter-cell pairs (RDCPs), in which both products of a single mitotic division are captured, allowing direct assessment of whether SCE events are reciprocal and copy-neutral or accompanied by structural alterations.

Here, we use sci-L3-Strand-seq to map Cas9-induced SCE at single-cell resolution across two experimental regimes: a single unique-locus cut and simultaneous cleavage at 237 repetitive targets. We show that a single Cas9 DSB at a unique site is a potent local trigger of SCE. At repetitive loci, on-target SCE enrichment is modest in bulk but becomes significant in a subset of cells with elevated recombination burden. RDCP analysis reveals that Cas9-induced SCEs are not uniformly benign: structural alterations at exchange junctions indicate that repair of two-ended, isochromatid Cas9 DSBs frequently deviates from error-free HR, in contrast to the clean reciprocal outcomes observed for spontaneous, replication-born SCEs in a companion study (Chovanec et al., 2026a). Wild-type Cas9 and nickase variants (D10A, H840A) are compared, and the consequences of disrupting DNA repair genes at the cut site are assessed. Together, these results provide the first genome-wide, single-cell map of Cas9-induced SCE and reveal that the isochromatid geometry of Cas9 breaks creates a distinct recombination landscape with both error-free and mutagenic outcomes.

Results

Cas9 duplex cleavage, but not nicking, can induce durable cell-cycle arrest at single and repetitive targets

To define how Cas9-induced lesions impact cell-cycle progression and SCE, we used recombinant Cas9 ribonucleoprotein (RNP) complexes (IDT), a widely adopted delivery format for genome editing that enables efficient generation of monoclonal knockouts as well as high-efficiency (>90%) polyclonal knockouts for bulk perturbation experiments. We compared four single-site sgRNA, each targeting a unique genomic locus, with a multi-target sgRNA recognizing 237 repetitive sites. Introduction of duplex-cleaving Cas9 at either a single site or across repetitive targets led to a pronounced G1/S arrest within 24 hours of delivery, indicating that even a single DSB is sufficient to trigger a robust checkpoint response, irrespective of whether the targeted locus resides in an essential gene (LIG3) or a non-essential gene (LIG4). Notably, this arrest persisted after Cas9 removal only in cells targeted with the repetitive sgRNA, consistent with sustained or cumulative DNA damage at multiple genomic sites (Figure 1A-B, Figure S1A).

Experimental design, cell cycle perturbation and genome-wide pileup of both spontaneous and Cas9-induced on-target SCE events.

A. Experimental design. BrdU was incorporated to enable SCE mapping by sci-L3-Strand-seq, either during the same 24-hour window as Cas9 RNP treatment or during a 24-hour period following Cas9 RNP washout. Wild-type Cas9 was used unless otherwise indicated for nickase variants. B. FACS analysis of cell-cycle progression following Cas9 RNP treatment. “24hrs” denotes concurrent Cas9 RNP treatment and BrdU labeling, whereas “24 ws BrdU” indicates BrdU labeling after Cas9 RNP washout. Each of the four single-targeting sgRNAs targets a distinct gene, as indicated. For the “LIG4 24hrs” condition, the small fraction of cells that successfully cycled into the subsequent G1 (6.71%) showed approximately half the Hoechst signal (centered at ∼50K) compared with the DAPI signal (centered at ∼100K), consistent with completion of cell division. C. Genome-wide pileup plots for both spontaneous and Cas9-induced on-target SCE events. The targeting site is indicated in the panel with the gene name and a vertical red bar. For sgRNAs targeting XRCC1 and XRCC4, SCEs from both the first division during RNP treatment and the post-washout division are shown. For LIG3 and LIG4, SCEs from the first division are shown. Additional pileups, including comparisons between breakpointR-based and rescued SCE calls, are provided in Figure S1B. D. Quantification of on-target SCE events for the four single-site sgRNAs in the first division (24 hrs) and in the subsequent division following RNP washout (24 ws).

Cas9 nickases, in which one of the two nuclease domains is inactivated (D10A or H840A), were developed to reduce off-target mutagenesis by converting DSBs into a single-strand nicks (Mali et al., 2013; Ran et al., 2013). Although paired nickases with offset guide RNAs can achieve on-target editing with improved specificity, individual nicks are not inherently inert, as they can be converted into DSBs by replication fork run-off or coincident nicking of the opposite strand. Given that only duplex cleavage at repetitive sequences induced a persistent cell cycle arrest after washout, we tested whether nickase-induced lesions at the same sites had a similar effect. As shown in Figure 1B, neither D10A nor H840A nickase variants measurably perturbed cell-cycle progression following washout. These findings indicate that duplex DNA cleavage, particularly when distributed across repetitive loci, imposes a durable cell-cycle barrier, whereas single-strand nicking is largely tolerated under these conditions.

A single Cas9 DSB induces potent local SCE

We next examined the genomic distribution of Cas9-induced and background spontaneous SCEs using genome-wide pileup analysis (Figure 1C, Figure S1B), in which we plot the total number of SCEs detected across all single cells within each 1 Mb window. Across all four sgRNAs targeting individual genomic loci, SCEs were strongly enriched at the programmed cut sites, demonstrating robust local induction of strand exchange. This enrichment was most pronounced in the first cell division following RNP delivery and was reduced 24 hours after washout. To account for low-coverage cells in which discrete breakpoints could not be confidently called, we performed a complementary analysis by quantifying strand-specific read imbalance within 10 Mb flanking the cut site, which similarly revealed clear evidence of SCE at the targeted loci. We define “on-target” events as the union of SCEs directly called by breakpointR and events recovered by this strand-state rescue approach. Across the single-site conditions, 30–92% of on-target events were directly called by breakpointR, typically >70%, with the exception of LIG4, whose proximity to the telomere makes breakpointR calling more difficult. The remaining on-target events were recovered by strand-state rescue, primarily in cells with lower per-cell sequencing depth. On-target enrichment remained robust when rescued calls were excluded (Figure S1B).

As summarized in Figure 1D, the combined on-target SCE rate reached up to 41% in the first division and 17% after washout. Overall, the on-target rate declined by approximately half following washout, with the largest reduction of 74%, indicating that Cas9-induced SCE is most efficiently captured immediately after cleavage and diminishes in subsequent cell cycles. Notably, while Cas9 RNP delivery achieves high bulk editing efficiencies (∼90%), this likely reflects multiple rounds of cleavage and repair at the same locus; consequently, the observed SCE frequencies should not be interpreted as a proxy for the final error-free HR or for knockout efficiency. Rather, the SCE rate reported here reflects the fraction of Cas9-induced breaks repaired through intersister exchange rather than cis repair on the same chromatid.

Cas9 targeting of DNA repair gene loci did not immediately alter overall SCE frequency per cell

Given the delayed functional loss following Cas9-mediated gene disruption, we do not expect spontaneous SCE to increase as a result of sgRNA-directed cutting within DNA repair genes over the timeframe of these experiments. Even biallelic frameshift mutations do not immediately eliminate the pre-existing protein pool, which must be diluted through turnover and cell division before a functional deficit can emerge. Consistent with this expectation, genome-wide SCE counts per cell were indistinguishable across all four gene-targeting conditions (LIG3, LIG4, XRCC1, XRCC4) at the 24-hour simultaneous labeling timepoint with all showing a median of 3 SCEs per cell, comparable to the wild-type with Cas9 cuts (Chovanec et al., 2026b) and Lig4 washout controls (pairwise Wilcoxon p > 0.09 for all 24-hour comparisons among single-site conditions, Figure S1C and Tab.S1). This contrasts with stable XRCC1 knockout, which increases spontaneous SCE approximately five-fold in our companion study (Chovanec et al., 2026a). One additional complication is that, although Cas9 RNP achieves >90% knockout efficiency in bulk assays and is therefore used as a substitute for siRNA, sorting BrdU-labeled cells in the subsequent G1 enriches for cells that escaped frameshift editing, particularly for essential genes; thus, 100% knockout in 90% of cells is not equivalent to 90% knockdown in every cell, representing a unique challenge for single-cell assays.

The LIG3 locus provides an informative example. As an essential gene, complete loss of LIG3 is expected to impair cell viability. In the 24-hour simultaneous labeling condition, >91% of recovered cells were in their first division following Cas9 delivery, and we observed no reduction in nuclei recovery, indicating minimal immediate selection. In contrast, in the washout condition, where labeling occurred in the second cycle, nuclei yield was markedly reduced, although the fraction of surviving cells progressing into the subsequent G1 was high. This pattern is consistent with selective loss of cells carrying biallelic loss-of-function alleles in the second cell cycle. The recovered cells, including those with on-target SCE at the LIG3 locus, likely retained functional LIG3, either through error-free repair (including SCE itself, which preserves sequence) or in-frame edits.

More broadly, although Cas9 RNP delivery achieves high bulk editing efficiencies by amplicon sequencing, this does not equate to functional equivalence with siRNA-mediated knockdown in single-cell assays. A population in which ∼90% of cells carry biallelic knockout is not equivalent to one in which all cells experience partial depletion. For essential genes in particular, single-cell readouts are inherently biased toward the surviving, functionally intact fraction, potentially obscuring the phenotype of interest.

On-target SCE enrichment at repetitive loci is mild but concentrated in high-SCE cells

We then asked whether SCE is enriched at the 237 repetitive target sites following Cas9 cleavage. To test this, we compared the number of SCE breakpoints falling within regions of interest (ROIs) flanking programmed cut sites against the number expected from size-matched random genomic intervals, using a Fisher’s exact test (Tab. 1). Across all cells, duplex Cas9 cleavage at 237 repetitive targets produced modest on-target SCE enrichment. The strongest signal came from the post-washout BrdU-labeled condition (“237cuts_24ws_BrdU”), which yielded a 1.20-fold enrichment (p = 4.89 x 10E-55, n = 2,547 cells), with an average of 6.13 on-target SCE breakpoints per cell. Adding Cas9 RNP and BrdU labeling simultaneously (“237cuts_24hrs”) showed a similar fold enrichment of 1.10 (p = 0.02, n = 214), with 5.72 on-target SCEs per cell. By contrast, the earlier WT237 experiment showed no significant enrichment (1.11-fold, p = 0.19, n = 74), likely reflecting the smaller sample size. The D10A nickase produced a marginally significant enrichment (1.09-fold, p = 0.01, n = 330), while the H840A nickase showed no enrichment (1.02-fold, p = 0.37, n = 757). These results indicate that duplex cleavage at repetitive loci induces detectable but mild on-target SCE enrichment in the bulk population, while single-strand nicking produces little or no measurable effect.

A stronger pattern emerged when analysis was restricted to cells with significantly elevated SCE count (>=9 SCE per cell). This threshold represents a 2-3 fold increase above the median of 3-4 SCE in non-cut wild-type cells (Chovanec et al., 2026b), and approximately two-fold above the median for the respective Cas9-treated populations, in which the median SCE per cell does not increase beyond 4 despite the introduction of up to 237 DSBs. In the post-washout BrdU condition, this high-SCE subset (n = 261 cells) showed 1.39-fold enrichment (p = 1.29 x 10E-41), with an average of 14.61 on-target SCEs per cell, compared to 6.13 in the full population. The simultaneous 24-hour condition, though limited to 4 cells, showed a comparable 1.39-fold enrichment with 11.50 on-target SCEs per cell. That the bulk median SCE remains unchanged despite 237 concurrent DSBs, while a minority of cells shows both elevated total SCE and concentrated on-target enrichment, suggests that most cells resolve Cas9-induced breaks through non-exchange, cis repair, and that inter-sister exchange is conditionally engaged in a SCE-permissive subpopulation. The WT237 (n = 8) and nickase conditions showed no significant enrichment in this subset, consistent with either insufficient statistical power or a genuine absence of on-target SCE in high-SCE cells following nicking. Relaxing the threshold to >=8 SCE per cell yielded qualitatively similar trends, particularly for the 24-hour condition (1.33-fold, p = 0.06, n = 10) and the post-washout condition (1.35-fold, p = 7.92 x 10E-45, n = 395).

Together, these data indicate that Cas9-induced SCE at repetitive targets is not uniformly distributed across the cell population. A subset of cells with elevated SCE burden shows both more on-target breaks (∼2-fold) and a higher fold enrichment of SCE at cut sites relative to random loci (1.39 vs 1.10-1.20 in the bulk), indicating that these cells not only sustain more Cas9 cleavage events but also resolve a greater fraction of them through inter-sister exchange. The high-SCE subset did not show evidence of lower sequencing quality, based on either sequencing coverage (p=0.13) or background SCE levels (p=0.17). However, we cannot distinguish whether the elevated on-target SCE reflects a cell-cycle state more permissive to both cutting and recombination, or stochastic variation in RNP uptake coupled with a recombination-prone chromatin environment.

RDCP analysis reveals large-scale SVs on chromosomes with induced SCE, as well as structural alterations at Cas9-induced SCE junctions

Reciprocal daughter cell pair (RDCP) analysis enables direct recovery of all four chromatids following replication, providing a stringent readout of SCE outcomes and associated structural changes. Leveraging the throughput of sci-L3-Strand-seq, we previously showed that spontaneous SCEs in XRCC1−/− cells are uniformly reciprocal and copy-neutral. This framework allows us to test whether Cas9 RNP-induced lesions similarly resolve without gross rearrangement. Given the limited sampling achievable with single targeting sites, we focused RDCP analysis on cells targeted with the 237-site sgRNA. We identified one RDCP under simultaneous labeling conditions, containing two reciprocal SCE events (Figure 2A). Following washout, we recovered 11 RDCPs, of which four exhibited associated structural variants (SVs) (Figure 2B). Across these four pairs, 14 of 26 events were accompanied by SVs, often arising across two consecutive cell divisions, whereas 12 events were reciprocal and copy-neutral. The remaining seven RDCPs comprised a total of 24 exclusively reciprocal, copy-neutral SCEs (Figure S2). Reconstruction of repair outcomes is detailed in Tab.S2. Our interpretation is further validated by haplotype-aware analysis in Figure S3. Overall, 14 of 50 SCE events were coupled with SVs, in contrast to 87 of 87 spontaneous SCE events in XRCC1−/− cells, which were uniformly reciprocal and copy-neutral in our companion study (Chovanec et al., 2026a). Additionally, we observed structural alterations at Cas9-induced SCE junctions consistent with the resolution of under-replicated regions (URR) or replication terminal zones as SCE (Figure 2B, Pair 4, Figure S3D). The mechanistic implications of these observations are explored further in the Discussion. We summarize RDCP counts, copy-neutral SCE counts, and SV-associated events across all conditions in Tab.S3.

RDCP analysis with sgRNA targeting 237 genomic sites.

In sci-L3-Strand-seq plots, reads mapping to the W and C template strands are plotted above and below zero on the y-axis, respectively. SCEs are identified as switches between W and C strand states along a chromosome. A. A single RDCP identified from 214 single cells following Cas9 RNP treatment targeting 237 sites with concurrent BrdU labeling. B. Four RDCPs exhibiting large-scale SVs recovered from cells labeled with BrdU after Cas9 RNP washout, corresponding to the second division.

Discussion

This study provides the first genome-wide, single-cell map of SCE induced by CRISPR/Cas9. Three principal findings emerge. First, a single Cas9 DSB at a unique genomic locus is a potent trigger of on-target SCE, with 17-41% of cells undergoing inter-sister strand switching within the same division as cleavage and BrdU labeling. When Cas9 is washed out and cells are labeled in the subsequent cell cycle, the on-target SCE rate drops by 50-74% to a range of 4-17%, confirming that the exchange signal reflects acute repair of the programmed, on-target DSB rather than a heritable alteration of SCE propensity. These strand switches represent non-local interactions between two sister chromatids that are invisible to amplicon sequencing and whole-genome sequencing, yet occur at frequencies comparable to those of the indel outcomes conventionally used to measure editing efficiency. Second, directing Cas9 to 237 repetitive targets produces only mild on-target SCE enrichment in the bulk population, but a subset of high-SCE cells shows more significant enrichment, consistent with a stochastic model in which a fraction of cells sustains a disproportionate share of recombinogenic cleavage events. Even in these high-SCE cells, the number of on-target SCEs per cell remains below 15, far fewer than the 237 available cut sites, indicating that most breaks never formed and/or were resolved by pathways that did not produce SCE. Third, Cas9 nickases (D10A and H840A) neither stall the cell cycle nor produce significant on-target SCE enrichment under the conditions tested, consistent with the expectation that single-strand nicks are less recombinogenic when delivered to a single genomic site without a paired offset nick, although we cannot exclude SCE formation at other time points.

Among the most informative outcomes in this dataset is a subset of Cas9-induced SCE events that display a distinctive structural signature in RDCP analysis: one daughter cell carries a strand switch with a WWC or WCC template-strand pattern at the SCE junction, while the paired daughter cell carries a deletion at the matching SCE junction (Figure 3). This “WWC-or- WCC/deletion pair” signature is inconsistent with classical homologous recombination, in which both daughters inherit reciprocal, copy-neutral strand switches. Instead, it resembles the products predicted by the URR resolution model, in which TRAIP-dependent unloading of the CMG helicase at stalled or converging replication forks exposes the template strands to nuclease cleavage, generating one-ended DSBs megabases apart, that are then joined by DNA polymerase theta to produce a sister chromatid with a deletion and an accompanying strand switch (Deng et al., 2019; Sonneville et al., 2019). Independently, van Vugt and colleagues demonstrated that SCEs can form in BRCA1/2-null cell lines, and are enriched at common fragile sites, providing cytogenetic evidence for an HR-independent, URR-coupled SCE pathway (Heijink et al., 2022). Because this pathway does not require end resection or strand invasion, it can operate in BRCA1/2-deficient cancer cells.

Haplotype-aware analysis of observed RDCP (Pair 4, chr1) shows SV patterns at the SCE junctions consistent with the predicted RDCP signature of SCE mediated by URRs or replication termination zones (green shaded area), although the lagging strands, rather than the leading strands, must be resolved to generate these mitotic breaks.

Oval depicts centromere. D1: daughter cell 1. D1 has a deletion region at the SCE junction. The corresponding region in D2 has a WC region within the haplotype with the SCE (WWC if depicted together with the other haplotype) as both strands are segregated without replication. Details in text.

Two recent studies revealed how the CDK1-TTF2-TRAIP axis is cell-cycle regulated to trigger mitotic CMG helicase disassembly and fork cleavage: one study showed a two-fold SCE reduction in mouse ES cells (Fujisawa and Labib, 2026), while the other showed that disrupting the TRAIP-TTF2 interaction reduced common fragile site deletions (Can et al., 2026). Our observation of the “WWC-or-WCC/deletion pair” signature in wild-type cells provides, to our knowledge, the first genetic evidence linking a deletion with SCE and revealing both W and C unreplicated template strands present in the reciprocal daughter cell, consistent with this mechanism at single-cell genomic resolution (illustrated in Figure 3), although this is limited by the observation of only one RDCP. Several mechanistic questions follow directly. Does TRAIP-mediated CMG unloading license structure-specific nucleases to process the exposed fork? Which resolvases (MUS81-EME1, SLX1-SLX4, GEN1) execute the cleavage? Note that this specific example actually represents lagging strand cleavage and is possible, although unlikely, due to Cas9 cleavage, RDCP analysis can resolve whether the leading or lagging strands are cleaved, which provides a unique opportunity if coupled with resolvase perturbations. Finally, we observe one such signature among approximately 36 otherwise normal SCEs in wild-type cells; determining the rate of this signature in BRCA1/2-null and other HR-perturbed conditions in spontaneous, rather than Cas9-induced SCE, will be essential for understanding the contribution of URR-type resolution to genome instability in cancer.

Limitation

A limitation of this study is that all experiments were performed in wild-type cells. In addition, the consequences of Cas9-induced DSBs may depend on both the cell-cycle stage at which cleavage occurs and the number of times a target is cut. Because Cas9 can undergo multiple rounds of cleavage and repair at the same target site, the genomic rearrangements observed here cannot necessarily be attributed to a single cleavage event, limiting mechanistic inference from individual repair outcomes. Approaches that provide tighter temporal control over Cas9 activity and the number of cleavage events will be important for resolving these questions in future studies. The genetic regulation of Cas9-induced SCE by the NHEJ, microhomology-mediated end joining (MMEJ), and HR pathways also remain to be dissected. In our companion paper, we found minimal role of NHEJ in regulating spontaneous SCE. However, it is unknown whether loss of NHEJ would redirect a greater fraction of Cas9 DSBs toward SCE, particularly given that Cas9 DSBs are two-ended, or whether loss of NHEJ abolishes apparent SCEs formed by dNHEJ across sister chromatids. Similarly, the role of MMEJ in competing with or contributing to SCE outcomes at isochromatid breaks and/or at URRs has not been tested. Systematic genetic perturbation in future experiments will be necessary to delineate the pathway hierarchy governing SCE at Cas9-induced isochromatid DSBs.

Conclusion

In summary, this work establishes that Cas9-induced DSBs trigger SCE at frequencies that are high for single-site cuts and detectable, though modest, for multi-target sites. Beyond quantifying on-target SCE rates, RDCP analysis reveals a “WWC-or-WCC/deletion pair” signature that cannot arise from classical homologous recombination and instead implicates a TRAIP/URR-type resolution pathway operating independently of BRCA1/2. This signature, observed here at single-cell genomic resolution in wild-type cells, provides a molecular framework for the HR-independent SCEs previously inferred from cytogenetic studies in BRCA-deficient backgrounds. More broadly, the isochromatid geometry of Cas9 breaks, in which both sisters are cleaved at the same locus, creates a recombination substrate structurally distinct from the single-ended, replication-born lesions that generate spontaneous SCE (Chovanec et al., 2026a), and this distinction is reflected in a measurable fraction of mutagenic outcomes. These findings underscore that single-cell, strand-level resolution is necessary for a complete accounting of the repair consequences of programmable nucleases.

Methods

Cell culture

BJ-5ta cells (CRL-4001, ATCC) were cultured in DMEM/F12 medium supplemented with 10% fetal bovine serum (FBS) and 1× penicillin-streptomycin at 37 °C in 5% CO2.

Cas9 RNP delivery

Cas9 ribonucleoprotein (RNP) complexes were assembled using Alt-R S. pyogenes Cas9 Nuclease V3 (1081058; IDT) or corresponding nickase variants (Alt-R Cas9 D10A Nickase V3; Alt-R Cas9 H840A Nickase V3) together with Alt-R tracrRNA (1072532; IDT) and Alt-R crRNA.

For single-targeting site experiments, crRNAs targeted coding exons of four DNA repair genes (LIG4: 5′-GCTTATACGGATGATCATAA-3′; XRCC4: 5′-ATGGTCATTCAGCATGGACT-3′; XRCC1: 5′-TGCAGGACACGACATGG-3′; LIG3: 5′-CTGTTAGGTACACATCACCG-3′). For the 237-cut condition, a crRNA targeting a repetitive element with 237 predicted genomic sites was used (5′-TCAGCACTTTGGGAGACCA-3′). CrRNA and tracrRNA were diluted to 1 μM in duplex buffer (11010301; IDT), annealed at 95 °C for 5 min, and cooled to room temperature for 30 min to form gRNA. Cas9 (or nickase) was diluted to 1 μM in OptiMEM, combined with gRNA and Cas9 PLUS reagent (CRISPRMAX kit, CMAX00003; Invitrogen), and incubated at room temperature for 5 min to assemble RNPs. Reverse transfection was performed with cells at 2×105 cells/mL.

The 237-site sgRNA targets a repetitive Alu sequence and was selected from a screen of >20,000 sgRNAs designed against repetitive sequences occurring at >200 genomic sites. The guide was selected for an intermediate cellular phenotype that permitted simultaneous targeting of a large number of genomic sites without excessive loss of cells; characterization of the broader repetitive-element screen is outside the scope of this study.

BrdU labeling conditions

Two labeling schemes were used: in the simultaneous condition (“24 hrs”), BrdU (40 μM final concentration) was added at the time of Cas9 RNP delivery, and cells were cultured for 24 hours prior to fixation, such that Cas9 cleavage and BrdU incorporation occurred within the same cell division. In the washout condition (“24 ws”), Cas9 RNP was delivered without BrdU for 24 hours, after which cells were washed and replated in fresh medium containing 40 μM BrdU for an additional 40 hours before fixation, labeling the division following Cas9 activity.

Sci-L3-Strand-seq library generation

Sci-L3-Strand-seq was performed as previously described (Chovanec and Yin, 2025; Yin et al., 2019). Briefly, cells were trypsinized and fixed in 1.5% formaldehyde in PBS at 1×106 cells/mL for 10 min at room temperature, followed by quenching with glycine. Cells were snap-frozen in nuclei freezing buffer (NFB: 50 mM Tris pH 8.0, 25% glycerol, 5 mM Mg(OAc)2, 0.1 mM EDTA, 5 mM DTT, 1× protease inhibitor).

Tagmentation was performed using Tn5 transposase with 24 first-round barcodes as described (Yin et al., 2019). After tagmentation (55 °C, 15 min), reactions were quenched with lysis buffer (LBT) and nuclei were pooled. Ligation with 72 second-round barcodes was followed by staining with Hoechst-33258 (10 ng/μL) and FACS sorting (200-300 nuclei per well into 96-well plates). Plates were stored at -80 °C.

Following gap extension, UV treatment (270 mJ/cm2, 365 nm), USER digestion, and UGI treatment, T7 in vitro transcription was performed (14-16 hours, 37 °C). RNA purification, reverse transcription, second-strand synthesis (third barcode incorporation), and Illumina library preparation were carried out using standard kits (NEB).

SCE analysis

Raw FASTQ files were processed using the sciL3Pipe Snakemake pipeline (v0.2.1), including barcode demultiplexing, alignment, and single-cell BAM generation. Breakpoint identification and filtering were performed using breakpointR (v1.10.0) and sciStrandR (v0.2.3). Cells passing quality filters (background estimate > 0 and ≤ 0.08; strand-neutral < 0.75; median reads per Mb > 2) were retained.

On-target SCE identification

SCEs were classified as on-target by two criteria. First, a breakpointR-called breakpoint within +/-3 Mb of the cut site was counted as a direct call. Second, a rescue procedure examined strand-state in two 10 Mb bins flanking the cut site (each offset 3 Mb from the gene boundary) and called a rescued SCE if the flanking bins showed a strand-state transition (e.g., WW-to-CC, WW-to-WC, CC-to-WC etc.). On-target strand switches were defined as the union of called and rescued events.

On-target SCE enrichment analysis (237-cut condition)

To test enrichment, observed SCEs overlapping cut-site windows were compared to those overlapping size-matched random genomic intervals using a one-sided’s exact test. Analyses were performed across all cells and repeated for high-SCE cells (cells with >=8 or >=9 post-filter SCEs).

Reciprocal daughter cell identification

The genome was partitioned into 1 Mb bins across autosomes and chromosome X. For each cell, W and C reads were counted per bin, and a strand ratio score was calculated as (W - C)/(W + C), yielding +1 (WW) or -1 (CC). Bins with fewer than four reads were set to 0. For each cell pair, a “sisterness” score was computed by averaging |scorei + scorej| across informative bins (|score| > 0.5 in at least one cell). Scores approach 0 for reciprocal strand inheritance and 2 for identical states. The top 20 candidate pairs with the lowest scores were selected and validated by visual inspection of Strand-seq ideograms.